---
title: "City R Training Series: Session 3"
subtitle: "Intro to Data Visualization"
output: html_notebook
---

```{r}
library("tidyverse")
library("readxl")
library("lubridate")
```

# Your First Plot

Consider this example, which should look familiar (most of it is from session 1).

```{r}

df <- read_excel("../data/311_sample.xlsx") %>%
  filter(Neighborhood %in% c("Middle East", "McElderry Park", "Milton-Montford")) %>%
  select(ServiceRequestNum, SRType, CreatedDate, CloseDate, Address)

```

If we want to count the number of service requests created in each year, we'll need to process CreatedDate again. Here's an alternative to parse_date_time. We specify the order the date/time components come in through the function name itself.

```{r}

plot_df <- df %>%
  mutate(Year = mdy_hms(CreatedDate))

```

Now that R knows that that column is a date, we can extract the year. 

```{r}

plot_df <- plot_df %>%
  mutate(Year = year(Year))

```

ggplot2 is a popular plotting package based on the "grammar of graphics" -- "a tool that enables us to concisely describe the components of a graphic." (Wickham 2010).

You can layer the following components to make a graphic:
* data and aesthetic mappings,
* geometric objects,
* scales,
* facet specification.
* statistical transformations, and
* the coordinate system.


```{r}

plot <- ggplot(plot_df, aes(x = Year)) +
  geom_bar(stat = "count")

plot

```

We can continue to modify this plot by layering more components.

```{r}

plot +
  geom_text(stat = "count", aes(label=..count..), vjust = -.5) +
  theme_minimal() +
  labs(title = "Middle East, McElderry Park, Milton-Montford Service Requests",
       x = "Calendar year",
       y = "Number of service requests")

```

# Dynamic vs Static Scripts

We saw this code in session 1. We first extracted this sample of 311 data in February 2019. What would we have to do in this example to change the dataset to a more updated data set? What if we want to show more rows?

```{r}

df <- read_excel("../data/311_sample.xlsx") %>%
  filter(Neighborhood %in% c("Middle East", "McElderry Park", "Milton-Montford")) %>%
  select(ServiceRequestNum, SRType, CreatedDate, CloseDate, Address) %>%
  arrange(desc(CreatedDate)) %>%
  mutate(Address = str_remove(Address, " MD,")) %>%
  write.csv("311_output_2019-02-25.csv")


```

As your scripts get more complex, there are more options that you need to sort through for any given dataset. 'Hard coding' in 'options' (such as file names) makes your code difficult to maintain over time. Pulling the options to the top helps you make changes quickly and efficiently.

```{r}
dir.create("../outputs/")
```

```{r}

file.input <- "../data/311_sample.xlsx"
file.output <- "../outputs/311_output_2019-02-25.csv"

df <- read_excel(file.input)

out <- df %>%
  filter(Neighborhood %in% c("Middle East", "McElderry Park", "Milton-Montford")) %>%
  select(ServiceRequestNum, SRType, CreatedDate, CloseDate, Address) %>%
  arrange(desc(CreatedDate)) %>%
  mutate(Address = str_remove(Address, " MD,"))

write.csv(out, file.output)
```

We can also make our scripts more dynamic in other ways:

```{r}

file.input <- "../data/311_sample.xlsx"

df <- read_excel(file.input)

out <- df %>%
  filter(Neighborhood %in% c("Middle East", "McElderry Park", "Milton-Montford")) %>%
  select(ServiceRequestNum, SRType, CreatedDate, CloseDate, Address) %>%
  arrange(desc(CreatedDate)) %>%
  mutate(Address = str_remove(Address, " MD,"))

write.csv(out, paste0("../outputs/311_output_", today(), ".csv"))

```


# Working with Excel Files

We often have to distribute data through Excel files. A lot of the tedium is in the formatting of the data. The openxlsx package can export formatted Excel sheets for you.

```{r}
install.packages("openxlsx")
```

```{r}
library("openxlsx")
excel <- createWorkbook()

# adding a tab
addWorksheet(excel, "Select Neighborhoods")

# choosing data for that tab
writeDataTable(excel, "Select Neighborhoods", out)

# exporting
saveWorkbook(excel, paste0("../outputs/311_output_", today(), ".xlsx"))
```

This gives us a very basically formatted spreadsheet. What options do we have?


```{r}
vignette("formatting", package = "openxlsx")
```

```{r}
?addWorksheet
```

```{r}
?writeDataTable
```

```{r}
excel <- createWorkbook()

# adding a tab
addWorksheet(excel, "Select Neighborhoods",
                    header = c("&[File]", "&[Tab]", "&[Date]"),
                    footer = c(NA, "&[Page] of &[Pages]", NA))

# choosing data for that tab
writeDataTable(excel, "Select Neighborhoods", out, tableStyle = "TableStyleLight1")

# auto setting column widths
setColWidths(excel, "Select Neighborhoods", 1:ncol(out), widths = "auto")

# setting the page orientation and ensuring that headers are repeated on subsequent pages when printing
pageSetup(excel, "Select Neighborhoods", "landscape", printTitleRows = 1)

# exporting
saveWorkbook(excel, paste0("../outputs/311_output_", today(), ".xlsx"), overwrite = TRUE)

```


